A White Noise Approach to Evolutionary Ecology

### 2020-04-09 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
improved proof of existence and uniqueness for the PDE

@@

Recall that $m(\nu,x)$ is shorthand for $m(K(\nu)(x,t),x)$. That is, $m:[0,+\infty)\times\mathbb{R}\to\mathbb{R}$. Following our assumptions of the main text, we have that $m(h,x)$ is differentiable with respect to both $x$ and $h$ and there exists $r\in\mathbb{R}$ such that $m(h,x)\leq r$ across all $h\geq0$ and $x\in\mathbb{R}$. As in the main text, we also assume the initial condition $\nu(x,0)$ satisfies
\begin{equation}
N(0)=\int_{\mathbb{R}}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{\mathbb{R}}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{\mathbb{R}}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $p(x,0)=\nu(x,0)/N(0)$, and we consider the PDE
\begin{equation}\label{PDE_SM}
\dot\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\Delta\nu(x,t).
\end{equation}

Replacing $m$ with it's upper bound $r\in\mathbb{R}$, equation (\ref{PDE_SM}) reduces to a simple parabolic equation that can be solved using standard techniques [@stanleyfarlow1993]. In particular, when $m(\nu,x)\equiv0$ denote the solution to (\ref{PDE_SM}) by $\nu_0(x,t)$. Then, denoting
\begin{equation}
\Phi(x,t)=\frac{\exp\left(-x^2/2\mu t\right)}{\sqrt{2\pi\mu t}},
\end{equation}
we have
\begin{equation}
\nu_0(x,t)=\int_{\mathbb{R}}\Phi(x-y,t)\nu(y,0)dy.
\end{equation}
In the more general case, when $m(\nu,x)\equiv r\in\mathbb{R}$, equation (\ref{PDE_SM}) has the solution $\nu_r(x,t)=e^{rt}\nu_0(x,t)$. Hence, $\nu_r(x,t)\geq0$ for all $x\in\mathbb{R}$ and $\int_\mathbb{R}\nu_r(x,t)dx=e^{rt}N(0)<+\infty$ for all $t\geq0$. Furthermore, denoting 
\begin{equation}
\begin{matrix}
N_r(t)&=&\int_\mathbb{R}\nu_r(x,t)dx, \\
p_r(x,t)&=&\nu_r(x,t)/N_r(t), \\
\bar x_r(t)&=&\int_\mathbb{R}xp_r(x,t)dx, \\
\sigma^2_r(t)&=&\int_\mathbb{R}(x-\bar x_r(t))^2p_r(x,t)dx, 
\end{matrix}
\end{equation}
we have
\begin{equation}
\bar x_r(t)=\int_\mathbb{R}x\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}yp(y,0)dy=\bar x(0),
\end{equation}
\begin{equation}
\sigma_r^2(t)=\int_\mathbb{R}(x-\bar x_r(t))^2\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}\Big((y-\bar x(0))^2+\mu t\Big)p(y,0)dy=\sigma^2(0)+\mu t.
\end{equation}
Hence, $|\bar x_r(t)|,\sigma_r^2(t)<+\infty$ for all $t\geq0$. For the sake of contradiction, suppose there exists $x\in\mathbb{R}$ and $t\geq0$ such that $\nu(x,t)>\nu_r(x,t)$. Then
\begin{equation}
\nu(x,t)-\nu(x,0)=\int_0^tm(\nu,x)\nu(x,s)+\frac{\mu}{2}\Delta\nu(x,s)ds>\int_0^tr\nu_r(x,s)+\frac{\mu}{2}\Delta\nu_r(x,s)ds=\nu_r(x,t)-\nu(x,0)
\end{equation}
which implies there exists $h\geq0$ and $x\in\mathbb{R}$ such that $m(h,x)>r$. But this contradicts our assumption $m(h,x)\leq r$ for all $h\geq0$ and $x\in\mathbb{R}$. So we have $\nu(x,t)\leq\nu_r(x,t)$ for each $x\in\mathbb{R}$ and $t\geq0$. This implies that $N(t)=\int_\mathbb{R}\nu(x,t)dx<+\infty$,
\begin{equation}
0\leq\int_\mathbb{R}x^2\nu(x,t)dx\leq\int_\mathbb{R}x^2\nu_r(x,t)dx<+\infty
\end{equation}
and in particular
\begin{equation}
0\leq\sigma^2(t)+\bar x^2(t)=\frac{1}{N(t)}\int_\mathbb{R}x^2\nu(x,t)dx<+\infty
\end{equation}
for each $t\geq0$.

## Regularity of deterministic solution \label{regular}

Here we show that solutions to the PDE described in the main text are continuous in time and space, twice continuously differentiable with respect to space and differentiable with respect to time. Our approach is an adaptation of that taken by Volpert [-@Volpert2014].

Recall that the PDE we consider is of the form
\begin{equation}\label{PDE_supp}
\dot\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\Delta\nu(x,t),
\end{equation}
where $m(\nu,x)$ bounded above by $r\in\mathbb{R}$ is shorthand for $m(K(\nu)(x,t),x)$ and $K(\nu)(x,t)=\int_\mathbb{R}\kappa(x,y)\nu(y,t)dy$ for some non-negative, bounded $\kappa$. Denoting $L^1(\mathbb{R})$ the Lebesgue space of integrable functions on $\mathbb{R}$ (ie, the set of functions $f:\mathbb{R}\to\mathbb{R}$ that satisfy $\|f\|_1=\int_\mathbb{R}|f(x)|dx<+\infty$ under the equivalence relation $f=g$ iff $\|f-g\|_1=0$), we assume $\nu(\cdot,0)\in L^1(\mathbb{R})$. Under these assumptions, we have already proved in SM \S\ref{finite} that $\nu(\cdot,t)\in L^1(\mathbb{R})$ for all $t\geq0$. Before proving the regularity of $\nu(x,t)$, we review the approach taken by Volpert [-@Volpert2014].

__Volperts approach__

In chapter 9 \S2.2 of Volpert [-@Volpert2014] complimentary conditions are assumed where $\kappa(x,y)=\varphi(x-y)$ for some integrable function $\varphi$ and initial condition $\nu(\cdot,0)\in L^\infty(\mathbb{R})$, where $L^\infty(\mathbb{R})$ is the Lebesgue space of bounded functions on $\mathbb{R}$ (ie, the set of functions $f:\mathbb{R}\to\mathbb{R}$ satisfying $\|f\|_\infty=\sup_{x\in\mathbb{R}}|f(x)|<+\infty$ under the equivalence relation $f=g$ iff $\|f-g\|_\infty=0$). Instead of the growth term $m(\nu,x)\nu(x,t)$ of PDE (\ref{PDE_supp}), Volpert worked with $f(\nu(x,t),K(\nu)(x,t))$ for some function $f:\mathbb{R}^2\to\mathbb{R}$ differentiable with respect to both arguments and satisfying, for some $q\geq0$,
\begin{equation}
f(x,y)\leq q(1+|x|+|y|), \ \forall \ x,y\in\mathbb{R}.
\end{equation}
Under these conditions, Volpert used semigroup theory to prove the solution $\nu(x,t)$ takes values in $L^\infty(\mathbb{R})$ for all $t\geq0$ and is differentiable with respect to time and twice continuously differentiable with respect to space. In particular, defining $\nu(t)$ as the function $x\to\nu(x,t)$ and keeping $\Phi(x,t)$ as defined in SM \S\ref{finite}, Volpert states the semigroup $\{S(t)\}_{t\geq0}$ given by 
\begin{equation}
S(t)\nu(0)=\int_0^t\frac{\mu}{2}\Delta\nu(x,s)ds=\int_\mathbb{R}\Phi(x-y,t)\nu(y,0)dy, \ \ \forall \ \ t\geq0, \ \nu(0)\in L^\infty(\mathbb{R})
\end{equation}
satisfies $\|S(t)\nu(0)\|_\infty\leq\|\nu(0)\|_\infty$ for all $t\geq0$ and $\nu(0)\in L^\infty(\mathbb{R})$. Then, by setting
\begin{equation}
F(v)=f\left(v(x),\int_\mathbb{R}\varphi(x-y)v(y)dy\right), \ \ \forall \ \ v\in L^\infty(\mathbb{R}),
\end{equation}
Volpert notes that $F$ is locally Lipschitz continuous on $L^\infty(\mathbb{R})$ and in particular $\|F(v)\|_\infty\leq q\|v\|_\infty$. Hence, under Volperts conditions, solutions to PDE (\ref{PDE_supp}) can be written as
\begin{equation}\label{volp_sol}
\nu(t)=S(t)\nu(0)+\int_0^tS(t-s)F(\nu(s))ds\in L^\infty(\mathbb{R}) \ \ \forall \ \ t\geq0, \ \nu(0)\in L^\infty(\mathbb{R}).
\end{equation}
Thanks to properties of the semigroup $\{S(t)\}_{t\geq0}$, solutions $\nu(t)$ given by (\ref{volp_sol}) satisfy the claimed regularity conditions.

__Adapting Volperts approach to our context__

To adapt Volperts approach to our situation, note that $\|S(t)\nu(0)\|_1=\|\nu(0)\|_1$ for all $\nu(0)\in L^1(\mathbb{R})$ and, setting
\begin{equation}
F(v)=m\left(\int_\mathbb{R}\kappa(x,y)v(y)dy,x\right)v(x), \ \ \forall \ \ v\in L^1(\mathbb{R}),
\end{equation}
we have $\|F(v)\|_1\leq r\|v\|_1$ for all $v\in L^1(\mathbb{R})$. Hence, using the same arguments employed by Volpert [-@Volpert2014], we have that solutions to PDE (\ref{PDE_supp}) under our assumptions can be written as
\begin{equation}
\nu(t)=S(t)\nu(0)+\int_0^tS(t-s)F(\nu(s))ds\in L^1(\mathbb{R}) \ \ \forall \ \ t\geq0, \ \nu(0)\in L^1(\mathbb{R})
\end{equation}
which therefore satisfy the claimed regularity conditions.

>>>>

## Existence, uniqueness and regularity of solution to deterministic PDE (\ref{eq1}) of the main text \label{regular}

Here we show that solutions to the PDE described in the main text exist, are unique and are twice continuously differentiable with respect to space. Our approach is an adaptation of that taken in chapter 9 \S2.2 of Volpert [-@Volpert2014].

Recall that the PDE we consider is of the form
\begin{equation}\label{PDE_supp}
\begin{matrix}
\dot\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\Delta\nu(x,t), \\
\\
\nu(x,0)=u(x),
\end{matrix}
\end{equation}
where $m(\nu,x)\leq r\in\mathbb{R}$ is shorthand for $m((K\nu)(x,t),x)$ and $(K\nu)(x,t)=\int_\mathbb{R}\kappa(x,y)\nu(y,t)dy$ for some non-negative, bounded $\kappa$. Denoting $C(\mathbb{R})$ the set of continuous functions on $\mathbb{R}$ and $L^1(\mathbb{R})$ the Lebesgue space of integrable functions on $\mathbb{R}$ (ie, the set of functions $f:\mathbb{R}\to\mathbb{R}$ that satisfy $\|f\|_1=\int_\mathbb{R}|f(x)|dx<+\infty$ under the equivalence relation $f=g$ iff $\|f-g\|_1=0$), we assume a continuous and integrable initial condition $\nu(\cdot,0)=u\in L^1(\mathbb{R})\cap C(\mathbb{R})$. We now prove the existence and uniqueness of solutions to this Cauchy problem.

Set 
\begin{equation}
\Phi(x,t)=\frac{\exp\left(-x^2/2\mu t\right)}{\sqrt{2\pi\mu t}}.
\end{equation}
Then the semigroup $\{S(t)\}_{t\geq0}$ given by 
\begin{equation}
S(t)v=\left\{\begin{matrix}
\int_0^t\frac{\mu}{2}\Delta\nu(x,s)ds=\int_\mathbb{R}\Phi(x-y,t)v(y)dy, & t>0 \\
 & \\
v, &t=0
\end{matrix}\right.
\end{equation}
for all $v\in L^\infty(\mathbb{R})$ satisfies $\|S(t)v\|_1=\|v\|_1$ for all $t\geq0$ and $v\in L^1(\mathbb{R})$. We also define the operator
\begin{equation}
(Fv)(x)=m\left(\int_\mathbb{R}\kappa(x,y)v(y)dy,x\right)v(x), \ \ \forall \ \ v\in L^1(\mathbb{R}).
\end{equation}
Thus $\|Fv\|_1\leq |r| \ \|v\|_1$ for all $v\in L^1(\mathbb{R})$ which implies $F$ is Lipschitz continuous. For $T>0$ set $L^{\infty,1}([0,T]\times\mathbb{R})$ as the set of functions $\nu:[0,T]\to L^1(\mathbb{R})$ that satisfy $\|\nu\|_{1,T}=\sup_{t\in[0,T]}\|\nu(t)\|_1<+\infty$ and consider the operator $G$ defined on $L^{\infty,1}([0,T]\times\mathbb{R})$ for fixed $u\in L^1(\mathbb{R})$ by
\begin{equation}
(Gv)(t)=S(t)u+\int_0^tS(t-s)(Fv(s))ds \ \ \forall \ \ t\in[0,T].
\end{equation}
Then for any $t\in[0,T]$ and $\nu_1,\nu_2\in L^{\infty,1}([0,T]\times\mathbb{R})$,

\begin{multline}
\|G\nu_1-G\nu_2\|_{1,T}=\sup_{t\in[0,T]}\left\|\int_0^tS(t-s)(F\nu_1(s)-F\nu_2(s))ds\right\|_1 \\ \leq \sup_{t\in[0,T]}t|r|\|\nu_1(t)-\nu_2(t)\|_1\leq T|r|\|\nu_1-\nu_2\|_{1,T}.
\end{multline}
Hence, when $T<\frac{1}{|r|}$ the operator $G$ is a contraction operator. In this case we apply Banach's fixed point theorem to demonstrate $G$ has a unique fixed point. That is, there exists a unique $\nu\in L^{\infty,1}([0,T]\times\mathbb{R})$ such that $G\nu=\nu$. Then, for $\nu=\nu(x,t)$ and considering $\nu(t)$ the function $x\to\nu(x,t)$, we have by definition $\nu$ is the mild solution of PDE (\ref{PDE_supp}) with initial condition $\nu(0)=u$. To extend this to arbitrarily large $T$ we take a similar approach as Evans [-@lawrenceevans2010] and choose $T_1<\frac{1}{|r|}$ and apply the above argument to the interval $[T_1,2T_1]$. Repeating this task a finite number of times shows $\nu$ is the mild solution of PDE (\ref{PDE_supp}) with initial condition $\nu(0)=u$ for arbitrarily large $T>0$. Applying lemma 3.3.2 of Henry [-@Henry1981] we have that the mild solution $\nu$ is in fact a strong solution to the Cauchy problem defined by PDE (\ref{PDE_supp}) with initial condition $\nu(0)=u$ for $t\in[0,T]$. In particular, following example 3.6 of Henry [-@Henry1981], this implies that $\nu(t)\in D(\Delta)$ for each $t\in(0,T)$ where $D(\Delta)=W^{2,2}(\mathbb{R})$ is the Sobolov space of square integrable functions with weak derivatives at least of order $2$. Then, according to a Sobolov embedding theorem [@Henry1981], we have $D(\Delta)\subset C^2(\mathbb{R})$ and hence solutions to the Cauchy problem (\ref{PDE_supp}) are twice continuously differentiable in their spatial parameter.

@@

Recall that $m(\nu,x)$ is shorthand for $m\big((K\nu)(x,t),x\big)$. That is, $m:[0,+\infty)\times\mathbb{R}\to\mathbb{R}$. Following our assumptions of the main text, we have that $m(h,x)$ is differentiable with respect to both $x$ and $h$ and there exists $r\in\mathbb{R}$ such that $m(h,x)\leq r$ across all $h\geq0$ and $x\in\mathbb{R}$. As in the main text, we also assume the initial condition $\nu(x,0)$ is continuous in $x$ and satisfies
\begin{equation}
N(0)=\int_{\mathbb{R}}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{\mathbb{R}}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{\mathbb{R}}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $p(x,0)=\nu(x,0)/N(0)$, and we consider the PDE
\begin{equation}\label{PDE_SM}
\dot\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\Delta\nu(x,t).
\end{equation}

In SM \S\ref{regular} we have shown PDE (\ref{PDE_SM}) with an integrable initial condition has a classical solution. Replacing $m$ with it's upper bound $r\in\mathbb{R}$, equation (\ref{PDE_SM}) reduces to a simple parabolic equation that can be solved using standard techniques [@stanleyfarlow1993]. In particular, when $m(\nu,x)\equiv0$ denote the solution to (\ref{PDE_SM}) by $\nu_0(x,t)$. Then, denoting
\begin{equation}
\Phi(x,t)=\frac{\exp\left(-x^2/2\mu t\right)}{\sqrt{2\pi\mu t}},
\end{equation}
we have
\begin{equation}
\nu_0(x,t)=\int_{\mathbb{R}}\Phi(x-y,t)\nu(y,0)dy.
\end{equation}
In the more general case, when $m(\nu,x)\equiv r\in\mathbb{R}$, equation (\ref{PDE_SM}) has the solution $\nu_r(x,t)=e^{rt}\nu_0(x,t)$. Hence, $\nu_r(x,t)\geq0$ for all $x\in\mathbb{R}$ and $\int_\mathbb{R}\nu_r(x,t)dx=e^{rt}N(0)<+\infty$ for all $t\geq0$. Furthermore, denoting 
\begin{equation}
\begin{matrix}
N_r(t)&=&\int_\mathbb{R}\nu_r(x,t)dx, \\
p_r(x,t)&=&\nu_r(x,t)/N_r(t), \\
\bar x_r(t)&=&\int_\mathbb{R}xp_r(x,t)dx, \\
\sigma^2_r(t)&=&\int_\mathbb{R}(x-\bar x_r(t))^2p_r(x,t)dx, 
\end{matrix}
\end{equation}
we have
\begin{equation}
\bar x_r(t)=\int_\mathbb{R}x\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}yp(y,0)dy=\bar x(0),
\end{equation}
\begin{equation}
\sigma_r^2(t)=\int_\mathbb{R}(x-\bar x_r(t))^2\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}\Big((y-\bar x(0))^2+\mu t\Big)p(y,0)dy=\sigma^2(0)+\mu t.
\end{equation}
Hence, $|\bar x_r(t)|,\sigma_r^2(t)<+\infty$ for all $t\geq0$. For the sake of contradiction, suppose there exists $x\in\mathbb{R}$ and $t\geq0$ such that $\nu(x,t)>\nu_r(x,t)$. Then
\begin{equation}
\nu(x,t)-\nu(x,0)=\int_0^tm(\nu,x)\nu(x,s)+\frac{\mu}{2}\Delta\nu(x,s)ds>\int_0^tr\nu_r(x,s)+\frac{\mu}{2}\Delta\nu_r(x,s)ds=\nu_r(x,t)-\nu(x,0)
\end{equation}
which implies there exists $h\geq0$ and $x\in\mathbb{R}$ such that $m(h,x)>r$. But this contradicts our assumption $m(h,x)\leq r$ for all $h\geq0$ and $x\in\mathbb{R}$. So we have $\nu(x,t)\leq\nu_r(x,t)$ for each $x\in\mathbb{R}$ and $t\geq0$. This implies that $N(t)=\int_\mathbb{R}\nu(x,t)dx<+\infty$,
\begin{equation}
0\leq\int_\mathbb{R}x^2\nu(x,t)dx\leq\int_\mathbb{R}x^2\nu_r(x,t)dx<+\infty
\end{equation}
and in particular
\begin{equation}
0\leq\sigma^2(t)+\bar x^2(t)=\frac{1}{N(t)}\int_\mathbb{R}x^2\nu(x,t)dx<+\infty
\end{equation}
for each $t\geq0$.
